dyngen: benchmarking with \textit{in silico} single cells

### 2019-10-16 content.tex
change cites

Here, synthetic data plays a crucial role in asserting minimum performance requirements for novel tools in anticipation of adequate real data. Generators of scRNA-seq data (e.g. splatter\cite{zappia_splattersimulationsinglecell_2017}, powsimR\cite{vieth_powsimrpoweranalysis_2017}, PROSSTT\cite{papadopoulos_prossttprobabilisticsimulation_2018} and SymSim\cite{zhang_simulatingmultiplefaceted_2019}) have already been widely used to explore the strengths and weakness of computational tools, both by method developers\cite{street_slingshotcelllineage_2018,parra_reconstructingcomplexlineage_2018,lummertzdarocha_reconstructioncomplexsinglecell_2018,lin_scclassifyhierarchicalclassification_2019} and independent benchmarkers\cite{duo_systematicperformanceevaluation_2018,saelens_comparisonsinglecelltrajectory_2019,soneson_biasrobustnessscalability_2018}.
However, a limitation of scRNA-seq profiles generators is that they would require significant methodological alterations to add additional modalities.

@@

Currently, \texttt{dyngen} focuses on simulating cells as standalone entities that are well mixed.
Splitting up the simulation space into separate subvolumes could pave the way to better study key cellular processes such as cell division, intercellular communication and migration\cite{smith_spatialstochasticintracellular_2019}.

% We developed \texttt{dyngen}, a simulator of single cells to benchmark and stress-test computational tools for single-cell omics.
%It supports a selection of existing experimental protocols, but can also simulate experiments for permitted by current technologies, such as generating omics profiles from the same cell at multiple timepoints.
%This data can help kick-start emerging domains with low data availability more safely by allowing method developers to test their method before numerous datasets become publicly available.
%\texttt{dyngen} has already been used successfully to evaluate trajectory inference\cite{saelens_comparisonsinglecelltrajectory_2019}, trajectory alignment\cite{vandenberge_trajectorybaseddifferentialexpression_2019}, and network inference\cite{pratapa_benchmarkingalgorithmsgene_2019} methods. Furthermore, we showed that it can also be used to evaluate differential network inference methods, and trajectory alignment methods. % TODO: And RNA velocity?

@@

One of the main processes involved in cellular dynamic processes is gene regulation, where regulatory cascades and feedback loops lead to progressive changes in expression and decision making. The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested \textit{in vivo}. One driver of bifurcation seems to be mutual antagonism, where two genes strongly repress each other\cite{rekhtman_directinteractionhematopoietic_1999,xu_regulationbifurcatingcell_2015}, forcing one of the two to become inactive \cite{graf_forcingcellschange_2009}. Such mutual antagonism can be modelled and simulated \cite{wang_quantifyingwaddingtonlandscape_2011, ferrell_bistabilitybifurcationswaddington_2012}. Although the two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into modules) repressing each other \cite{yosef_dynamicregulatorynetwork_2013}.

@@

\paragraph{Step 3, sampling the target subnetwork.} 
A user-defined number of target genes are added to the GRN. Target genes regulated by a TF or another target gene, but is always downstream of at least one TF. To sample the interactions between target genes, one of the many FANTOM5\cite{lizio_gatewaysfantom5promoter_2015} GRNs is sampled. The currently existing TFs are mapped to regulators in the FANTOM5 GRN. The targets are drawn from the FANTOM5 GRN, weighted by their page rank value. For each target, at most $x$ regulators are sampled from the induced FANTOM5 GRN (default $x=5$). The interactions connecting a target gene and its regulators are added the GRN.

@@

\texttt{dyngen} uses Gillespie's stochastic simulation algorithm (SSA)\cite{gillespie_exactstochasticsimulation_1977} to simulate dynamic processes. An SSA simulation is an iterative process where at each iteration one reaction is triggered.

@@

\subsubsection{Sample molecules} 
Molecules are sampled from the simulation to replicate how molecules are experimentally sampled. A real dataset is downloaded from a repository of single-cell RNA-seq datasets\cite{cannoodt_singlecellomicsdatasets_2018}. For each \textit{in silico} cell $i$, draw its library size $ls_i$from the distribution of transcript counts per cell in the real dataset. The capture rate $cr_j$ of each \textit{in silico} molecule type $j$ is drawn from $N(1, 0.05)$. Finally, for each cell $i$, draw $ls_i$ molecules from the multinomial distribution with probabilities $cr_j \times ab_{i,j}$ with $ab_{i,j}$ the molecule abundance level of molecule $j$ in cell $i$.

>>>>

Here, synthetic data plays a crucial role in asserting minimum performance requirements for novel tools in anticipation of adequate real data. Generators of scRNA-seq data (e.g. splatter \cite{zappia_splattersimulationsinglecell_2017}, powsimR \cite{vieth_powsimrpoweranalysis_2017}, PROSSTT \cite{papadopoulos_prossttprobabilisticsimulation_2018} and SymSim \cite{zhang_simulatingmultiplefaceted_2019}) have already been widely used to explore the strengths and weakness of computational tools, both by method developers \cite{street_slingshotcelllineage_2018,parra_reconstructingcomplexlineage_2018,lummertzdarocha_reconstructioncomplexsinglecell_2018,lin_scclassifyhierarchicalclassification_2019} and independent benchmarkers \cite{duo_systematicperformanceevaluation_2018,saelens_comparisonsinglecelltrajectory_2019,soneson_biasrobustnessscalability_2018}.
However, a limitation of scRNA-seq profiles generators is that they would require significant methodological alterations to add additional modalities.

@@

Currently, \texttt{dyngen} focuses on simulating cells as standalone entities that are well mixed.
Splitting up the simulation space into separate subvolumes could pave the way to better study key cellular processes such as cell division, intercellular communication and migration \cite{smith_spatialstochasticintracellular_2019}.

% We developed \texttt{dyngen}, a simulator of single cells to benchmark and stress-test computational tools for single-cell omics.
%It supports a selection of existing experimental protocols, but can also simulate experiments for permitted by current technologies, such as generating omics profiles from the same cell at multiple timepoints.
%This data can help kick-start emerging domains with low data availability more safely by allowing method developers to test their method before numerous datasets become publicly available.
%\texttt{dyngen} has already been used successfully to evaluate trajectory inference \cite{saelens_comparisonsinglecelltrajectory_2019}, trajectory alignment \cite{vandenberge_trajectorybaseddifferentialexpression_2019}, and network inference \cite{pratapa_benchmarkingalgorithmsgene_2019} methods. Furthermore, we showed that it can also be used to evaluate differential network inference methods, and trajectory alignment methods. % TODO: And RNA velocity?

@@

One of the main processes involved in cellular dynamic processes is gene regulation, where regulatory cascades and feedback loops lead to progressive changes in expression and decision making. The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested \textit{in vivo}. One driver of bifurcation seems to be mutual antagonism, where two genes strongly repress each other \cite{rekhtman_directinteractionhematopoietic_1999,xu_regulationbifurcatingcell_2015}, forcing one of the two to become inactive \cite{graf_forcingcellschange_2009}. Such mutual antagonism can be modelled and simulated \cite{wang_quantifyingwaddingtonlandscape_2011, ferrell_bistabilitybifurcationswaddington_2012}. Although the two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into modules) repressing each other \cite{yosef_dynamicregulatorynetwork_2013}.

@@

\paragraph{Step 3, sampling the target subnetwork.} 
A user-defined number of target genes are added to the GRN. Target genes regulated by a TF or another target gene, but is always downstream of at least one TF. To sample the interactions between target genes, one of the many FANTOM5 \cite{lizio_gatewaysfantom5promoter_2015} GRNs is sampled. The currently existing TFs are mapped to regulators in the FANTOM5 GRN. The targets are drawn from the FANTOM5 GRN, weighted by their page rank value. For each target, at most $x$ regulators are sampled from the induced FANTOM5 GRN (default $x=5$). The interactions connecting a target gene and its regulators are added the GRN.

@@

\texttt{dyngen} uses Gillespie's stochastic simulation algorithm (SSA) \cite{gillespie_exactstochasticsimulation_1977} to simulate dynamic processes. An SSA simulation is an iterative process where at each iteration one reaction is triggered.

@@

\subsubsection{Sample molecules} 
Molecules are sampled from the simulation to replicate how molecules are experimentally sampled. A real dataset is downloaded from a repository of single-cell RNA-seq datasets \cite{cannoodt_singlecellomicsdatasets_2018}. For each \textit{in silico} cell $i$, draw its library size $ls_i$from the distribution of transcript counts per cell in the real dataset. The capture rate $cr_j$ of each \textit{in silico} molecule type $j$ is drawn from $N(1, 0.05)$. Finally, for each cell $i$, draw $ls_i$ molecules from the multinomial distribution with probabilities $cr_j \times ab_{i,j}$ with $ab_{i,j}$ the molecule abundance level of molecule $j$ in cell $i$.
